Linear response theory and optimal control for a molecular system under nonequilibrium conditions

### 2013-08-22 writedown/linear.response/paper.tex
new revise by han, send to C and C.

Standard molecular dynamics simulations are dealing with systems in thermal equilibrium; in this case they are tuned to the canonial or Boltzmann distribution in the sense that either (1) if one starts from this distribution it remains invariant under the dynamics or (2) if one generates a very long trajectory it samples state space with respect this distribution, that is, every possible state of the molecular system under consideration is visited according to the probability given by it.  Obviously, this allows to compute equilibrium expectation values with respect to the  canonical distribution simply be computing long trajectories.

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the nonequilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations; see, e.g., \cite{marconi2008,ruelle2009} and the references therein. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. Linear response formulae on the infinite time-horizon have been shown to hold under rather mild assumptions even for nonequilibrium systems that possess a unique invariant measure \cite{hairer2008}.

@@

The outline of the article is as follows: First, we will review the derivation of finite time-horizon linear response formulae for general diffusion processes and diffusions of Langevin type. Next, we will show how to derive first and second order response formulas for genuine nonequilibrium systems and how to apply these formula to the computation of optimal controls. Finally, we will validate the nonequilibrium linear response formula and its use for optimal control for simple test cases. This numerical experiments will also outline that the use of the linear response formula is imperative for numerical efficiency and allows to extend the applicability of linear response theory to stronger perturbations.

@@

\subsubsection*{Girsanov transformation}
We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. To this end, let $x_{t}=x_{t}(\omega)$ and $X_{t}=X_{t}(\omega)$ be the solutions of our generic stochastic differential equations
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t}\,,\quad 0\le t\le T  \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad 0\le t\le T \label{sde2-2}
\end{align}
\end{subequations}
for $T<\infty$ and deterministic initial conditions
\[
x_{0}(\omega) = X_{0}(\omega) = x\quad\textrm{(almost surely)}\,.
\]
Suppose that there exists an auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to rewrite (\ref{sde2-2}) as
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $P$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $Q$ on the space of continuous trajectories by 
\[
dQ=M_{T}dP
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any functional 
\[
\varphi(\omega)=\int_{0}^{T}f((X_{s}(\omega))\,ds + g(X_{T}(\omega))
\] 
that is integrable \wrt $\nu$, we have the identity\footnote{One of the omitted technical details is Novikov's condition \cite[pp.~162]{oksendal2003stochastic} that guarantees that $(M_{t})_{0\le t\le T}$ is a Martingale, which by $\bE_{Q}[1]=\bE_{P}[M_{T}]=1$ implies normalizability of the new probability measure $Q$.}
\[
\bE_{P}[\varphi] := \int_{\Omega} \varphi(\omega)\, dP(\omega) = \int_{\Omega} \varphi(\omega) M_{T}^{-1}(\omega)\,dQ(\omega) =: \bE_{Q}[M^{-1}_{T}\varphi]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $P$ relative to $Q$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $Q$, and the expectation is over all realizations of (\ref{sde2-3}) starting from $X_{0}=x$. On the other hand, $X_{t}$ under $P$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}).

@@

Linearization of $M_{T}^{-1}$ about $\eps=0$, assuming that the control has bounded variance, yields the alternative linear response formula 
\begin{equation}\label{lr-alt}
\lim_{\eps\to 0}\frac{\bE_{P_{x}^{\eps}}[\varphi] + \bE_{P_{x}^{0}}[\varphi]}{\eps} = \bE_{P_{x}^{0}}\left[\varphi \int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\,,
\end{equation}
where, in analogy with the previous linear response formula, $P_{x}^{\eps}$ and $P_{x}^{0}$ denote the path probability measures of perturbed and unperturbed paths---i.e.~the trajectories obtained from (\ref{sde2-2}) and (\ref{sde2-1})---with deterministic initial condition. The path measure $P^{\eps}_{x}$ should not be confused with the phase space density $\rho^{\eps}(x,t)$ in Section \ref{ssec:equiLRT} that solves the Fokker Planck equation. Note moreover that $\varphi$ on the right hand side of (\ref{lr-alt}) is understood as a functional of $(x_{t})_{0\le t\le T}$, i.e.~the solution to (\ref{sde2-1}) with initial condition $x_{0}=x$, with $\xi_{t}$ being the solution to $a(x_{t})\xi_{t}=v_{t}$.

@@

We are interested in the dynamics under a small perturbation $\delta u$ of the control $u$, Specifically, we assume that $v$ in (\ref{sde2}) is of the form 
\begin{equation}\label{eqn:langevin-1c}
v_{t} = c(X_{t})\delta u(t)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sqrt{2\beta^{-1}}R^{1/2} \xi_{t} = D(X_{t})\,\delta u(t)\,.
\]
Hence the equation for $\xi_{t}$ is solvable if and only if $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(R^{1/2})$, which, since $R$ has full rank, means that $D(\cdot)$ must be invertible almost everywhere.

@@

The nonequilibrium linear response formula can be used for solving certain optimal control problems. To this end, let us remain in the setting of equations (\ref{eqn:langevin-1d})--(\ref{eqn:neq-response}) and assume that we are interested choosing the nonequilibrium forcing $u$, such that the expected value $I=\bE[\psi]$ of some utility function 
\[
\psi(u) = \int_{0}^{T}\left\{f(x_{s}) - \frac{1}{2}|u(s)|^{2}\right\}\,dt + g(x_{T}) 
\]
is maximized where $x_{t}$ is the solution to the controlled Langevin equation (\ref{eqn:langevin-1d}) for $\eps=0$; the functions $f$ and $g$ are the running cost and the terminal cost, which are assumed to be continuous and bounded from above; without loss of generality, $f$ and $g$ are assumed to depend only on the positions $q:=x_{1}$ and not on the momenta $p:=x_{2}$. The quadratic term is a penalization that makes sure that the optimal control remains bounded \cite{schuette2012,hartmann2012}. Let us moreover assume that the controls are open-loop (i.e. without feedback) and can be represented by 
\[
u(t) = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\]
with suitably chosen time dependent, bounded and Lipschitz continuous vector fields $\Phi_k\colon[0,T]\to\R^{n}$, $k=1,\ldots,K$. %In principle we could even go for closed-loop (i.e.~feedback) controls by making $\Phi_{k}$ state dependent. 
Specifically, we want to solve the optimal control problem of the following standard form: 
\begin{equation}\label{ocp1}
\begin{aligned}
 \max_{u\in\cU} \,& I(u)\quad \textrm{s.t.}\\
dq_t & =  \nabla_p H(q_t,p_t)dt\\
dp_t & =  -\nabla_q H(q_t,p_t)dt - \gamma \nabla_p H(q_t,p_t)dt + D(q_t)u(t) dt+\sigma dB_t\,.
\end{aligned}
\end{equation}
where 
\begin{equation}\label{ocp2}
I(u) = \bE\left[\int_{0}^{T}\left\{f(q_{s}) - \frac{1}{2}|u(s)|^{2}\right\}\,dt + g(q_{T}) \right]\,,
\end{equation}
with the expectation being over all realizations of the controlled Langevin dynamics in (\ref{ocp1}) with any prescribed initial distribution. Here, the space of admissible controls $\cU$ consists of all bounded controls of the form 
\begin{equation}\label{admissible}
u(t) = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\end{equation}

@@

In principle, optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i.e.~by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Expect for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,,
\end{equation*} 
For existence of the functional derivative \wrt general It\^o stochastic processes, we refer to \cite{bell2006}.  
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate where $\delta a_{1},\ldots,\delta a_{K}$ are the coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{P^{\eps}}[\psi(u+\eps \delta u)] - \bE_{P^{0}}[\psi(u)]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{P^0}\left[\psi\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} + \int_{0}^{T}u(s)\cdot\Phi_{k}(s)\,ds \right] \,
\end{equation}
and $\bE_{P^0}[\cdot]$ being the expectation over all realizations of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate and arbitrary initial conditions $x=(q,p)$, e.g., 
\[
\bE_{P^{0}}[\psi] = \int_{\R^{2n}}\bE_{P^{0}_{x}}[\psi]\,d\mu_{\infty}(x) \,,
\]
with $P_{x}^{0}$ denoting the corresponding path measure with deterministic initial condition. Note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state and the time elapsed since $t=0$.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/fig-split-pot.eps}
  \caption{The single-well potential with  splitting driving
    force. Time $T = 20$ ps.  Since the driving force is
    of gradient form, $ D( q) = -\nabla_{q}V( q)$,
    we plot the nonequilibrium driving energy of the system.  The red,
    green and blue lines are the potential energy without
    nonequilibrium driving, with nonequilibrium driving and
    the nonequilibrium driving perturbed by $\eps\delta u(t)$,
    respectively.  }
  \label{fig:tmp4}
\end{figure}

We use the idea of nonequilibrium linear response to investigate the
nonequilibrium phase space probability density, denoted by $\rho^\eps(q, p, t)$, of a one-dimensional model system: one particle in a
splitting single-well potential as shown in Fig.~\ref{fig:tmp4}.  For convenience, we let the mass of
the particle to be 1~\textsf{amu}, and the friction coefficient to be
$1\,\textsf{ps}^{-1}$.
The temperature is the
room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
The unperturbed Hamiltonian of the system is
given by:
\begin{align}
  H ( p,  q) = \frac 12  p^2 + U( q) 
\end{align}
with potential
\begin{align}
  U( q) = \frac12\,k\, q^2 
\end{align}
Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^2)$.
See the red line in Fig.~\ref{fig:tmp4} for the  potential $U$.
%$\textsf{kJ} / (\textsf{mol nm}^4)}$
The nonequilibrium driving $ D$ is given by a force of gradient form:
\begin{align}
   D( q) = -\nabla_{ q} V( q) ,
\end{align}
where the driving potential $V( q)$ has a Gaussian profile:
\begin{align}
  V( q) = \frac{1}{\sqrt{2\pi \sigma^2}}
  \exp\Big\{-\frac{ q^2}{2\sigma^2}\Big\}
\end{align}
we use $\sigma = 0.16$~\textsf{nm}.  The strength of nonequilibrium
driving $u(t)$ is set to be linearly growing, i.e. $ u(t) = k_e\cdot t/T$,
where $k_e$ is a unitless constant.
We
consider the perturbation to the system given by $\eps\delta u(t) = \eps k_e \cdot t/T$.
We consider the following parameters: end time $T = 20$~\textsf{ps}, 
$k_e = 1$ and $\eps = 1$, see Fig.~\ref{fig:tmp4}
for the nonequilibrium driving potential and perturbed
potential at time $t = T$. The initial distribution $\rho^0(q, p, 0)$ is set to
be equilibrium distribution of the unperturbed system.
\recheck{
  We use Euler scheme with time step of $10^{-4}$ to discretize the
  Langevin equation. The initial configurations for nonequilibrium
  simulation are generated by taking configurations
  from an equilibrium simulation of $10^6$~\textsf{ps} at an 1~\textsf{ps} time interval.
  Then starting with these initial configurations according to $\rho_{\infty}\propto\exp(-\beta H)$,  the nonequilibrium
  Langevin equation is integrated until 20~\textsf{ps} with the same numerical
  scheme and time step.
}

>>>>

Standard molecular dynamics simulations are dealing with systems in thermal equilibrium; in this case they are tuned to the canonical or Boltzmann distribution in the sense that either (1) if one starts from this distribution it remains invariant under the dynamics or (2) if one generates a very long trajectory it samples state space with respect this distribution, that is, every possible state of the molecular system under consideration is visited according to the probability given by it.  Obviously, this allows to compute equilibrium expectation values with respect to the  canonical distribution simply be computing long trajectories.

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the nonequilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations; see, e.g., \cite{marconi2008,ruelle2009} and the references therein. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. Linear response formulas on the infinite time-horizon have been shown to hold under rather mild assumptions even for nonequilibrium systems that possess a unique invariant measure \cite{hairer2008}.

@@

The outline of the article is as follows: First, we will review the derivation of finite time-horizon linear response formulas for general diffusion processes and diffusions of Langevin type. Next, we will show how to derive first and second order response formulas for genuine nonequilibrium systems and how to apply these formula to the computation of optimal controls. Finally, we will validate the nonequilibrium linear response formula and its use for optimal control for simple test cases. This numerical experiments will also outline that the use of the linear response formula is imperative for numerical efficiency and allows to extend the applicability of linear response theory to stronger perturbations.

@@

\subsubsection*{Girsanov transformation}
We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. To this end, let $x_{t}=x_{t}(\omega)$ and $X_{t}=X_{t}(\omega)$ be the solutions of our generic stochastic differential equations
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t}\,,\quad 0\le t\le T  \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad 0\le t\le T \label{sde2-2}
\end{align}
\end{subequations}
for $T<\infty$ and deterministic initial conditions
\[
x_{0}(\omega) = X_{0}(\omega) = x\quad\textrm{(almost surely)}\,.
\]
Suppose that there exists an auxiliary stochastic process \redc{ $\xi_{t}\in\R^{n}$} such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to rewrite (\ref{sde2-2}) as
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $P$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $Q$ on the space of continuous trajectories by 
\[
dQ=M_{T}dP
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any functional 
\[
\varphi(\omega)=\int_{0}^{T}f((X_{s}(\omega))\,ds + g(X_{T}(\omega))
\] 
that is integrable \wrt $\nu$, we have the identity\footnote{One of the omitted technical details is Novikov's condition \cite[pp.~162]{oksendal2003stochastic} that guarantees that $(M_{t})_{0\le t\le T}$ is a Martingale, which by $\bE_{Q}[1]=\bE_{P}[M_{T}]=1$ implies normalizability of the new probability measure $Q$.}
\[
\bE_{P}[\varphi] := \int_{\Omega} \varphi(\omega)\, dP(\omega) = \int_{\Omega} \varphi(\omega) M_{T}^{-1}(\omega)\,dQ(\omega) =: \bE_{Q}[M^{-1}_{T}\varphi]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $P$ relative to $Q$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $Q$, and the expectation is over all realizations of (\ref{sde2-3}) starting from $X_{0}=x$. On the other hand, $X_{t}$ under $P$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}).

@@

Linearization of $M_{T}^{-1}$ about $\eps=0$, assuming that the control has bounded variance, yields the alternative linear response formula 
\begin{equation}\label{lr-alt}
\lim_{\eps\to 0}\frac{\bE_{P_{x}^{\eps}}[\varphi] - \bE_{P_{x}^{0}}[\varphi]}{\eps} = \bE_{P_{x}^{0}}\left[\varphi \int_{0}^{T}\xi_{s}\cdot dB_{s} \right]\,,
\end{equation}
where, in analogy with the previous linear response formula, $P_{x}^{\eps}$ and $P_{x}^{0}$ denote the path probability measures of perturbed and unperturbed paths---i.e.~the trajectories obtained from (\ref{sde2-2}) and (\ref{sde2-1})---with deterministic initial condition. The path measure $P^{\eps}_{x}$ should not be confused with the phase space density $\rho^{\eps}(x,t)$ in Section \ref{ssec:equiLRT} that solves the Fokker Planck equation. Note moreover that $\varphi$ on the right hand side of (\ref{lr-alt}) is understood as a functional of $(x_{t})_{0\le t\le T}$, i.e.~the solution to (\ref{sde2-1}) with initial condition $x_{0}=x$, with $\xi_{t}$ being the solution to $a(x_{t})\xi_{t}=v_{t}$.

@@

We are interested in the dynamics under a small perturbation $\delta u$ of the control $u$, Specifically, we assume that $v$ in (\ref{sde2}) is of the form 
\begin{equation}\label{eqn:langevin-1c}
v_{t} = c(X_{t})\delta u(t)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sqrt{2\beta^{-1}}\recheck{\gamma}^{1/2} \xi_{t} = D(X_{t})\,\delta u(t)\,.
\]
\recheck{Hence the equation for $\xi_{t}$ is solvable if and only if $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(\gamma^{1/2})$, which, since $\gamma$ has full rank, means no restriction on $D(\cdot)$.}
  % means that $D(\cdot)$ must be invertible almost everywhere.}

@@

The nonequilibrium linear response formula can be used for solving certain optimal control problems. To this end, let us remain in the setting of equations (\ref{eqn:langevin-1d})--(\ref{eqn:neq-response}) and assume that we are interested in choosing the nonequilibrium forcing $u$, such that the expected value $I=\bE[\psi]$ of some utility function 
\[
\psi(u) = \int_{0}^{T}\left\{f(x_{s}) - \frac{1}{2}|u(s)|^{2}\right\}\,dt + g(x_{T}) 
\]
is maximized where $x_{t}$ is the solution to the controlled Langevin equation (\ref{eqn:langevin-1d}) for $\eps=0$; the functions $f$ and $g$ are the running cost and the terminal cost, which are assumed to be continuous and bounded from above; without loss of generality, $f$ and $g$ are assumed to depend only on the positions $q:=x_{1}$ and not on the momenta $p:=x_{2}$. The quadratic term is a penalization that makes sure that the optimal control remains bounded \cite{schuette2012,hartmann2012}. Let us moreover assume that the controls are open-loop (i.e. without feedback) and can be represented by 
\[
u(t) = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\]
with suitably chosen time dependent, bounded and Lipschitz continuous vector fields $\Phi_k\colon[0,T]\to\R^{n}$, $k=1,\ldots,K$. %In principle we could even go for closed-loop (i.e.~feedback) controls by making $\Phi_{k}$ state dependent. 
Specifically, we want to solve the optimal control problem of the following standard form: 
\begin{equation}\label{ocp1}
\begin{aligned}
 \max_{u\in\cU} \,& I(u)\quad \textrm{s.t.}\\
dq_t & =  \nabla_p H(q_t,p_t)dt\\
dp_t & =  -\nabla_q H(q_t,p_t)dt - \gamma \nabla_p H(q_t,p_t)dt + D(q_t)u(t) dt+\sigma dB_t\,,
\end{aligned}
\end{equation}
\redc{with $(q_0, p_0)$ being subject to a given initial distribution $\mu$, and}
\begin{equation}\label{ocp2}
I(u) = \bE\left[\int_{0}^{T}\left\{f(q_{s}) - \frac{1}{2}|u(s)|^{2}\right\}\,dt + g(q_{T}) \right]\,,
\end{equation}
with the expectation being over all realizations of the controlled Langevin dynamics in (\ref{ocp1}) with \recheck{the} prescribed initial distribution. Here, the space of admissible controls $\cU$ consists of all bounded controls of the form 
\begin{equation}\label{admissible}
u(t) = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\end{equation}

@@

In principle, optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i.e.~by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Except for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can do a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,.
\end{equation*} 
For existence of the functional derivative \wrt general It\^o stochastic processes, we refer to \cite{bell2006}.  
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate where $\delta a_{1},\ldots,\delta a_{K}$ are the coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{P^{\eps}}[\psi(u+\eps \delta u)] - \bE_{P^{0}}[\psi(u)]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{P^0}\left[\psi\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} + \int_{0}^{T}u(s)\cdot\Phi_{k}(s)\,ds \right] \,,
\end{equation}
and $\bE_{P^0}[\cdot]$ being the expectation over all realizations of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate and arbitrary initial conditions $x=(q,p)$, e.g., 
\[
\bE_{P^{0}}[\psi] = \int_{\R^{2n}}\bE_{P^{0}_{x}}[\psi]\,d\recheck{\mu}(x) \,,
\]
with $P_{x}^{0}$ denoting the corresponding path measure with deterministic initial condition,
\redc{and $\mu$ is the prescribed initial distribution.}
Note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state and the time elapsed since $t=0$.

@@

\begin{figure}
  \centering
  % \includegraphics[width=0.4\textwidth]{figs/fig-split-pot.eps}
  \includegraphics[width=0.5\textwidth]{figs/fig-split-pot.eps}
  \caption{The single-well potential with  splitting driving
    force. Time $T = 20$ ps.  Since the driving force is
    of gradient form, $ D( q) = -\nabla_{q}V( q)$,
    we plot the nonequilibrium driving energy of the system.  The red,
    green and blue lines are the potential energy without
    nonequilibrium driving, with nonequilibrium driving and
    the nonequilibrium driving perturbed by $\eps\delta u(t)$,
    respectively.  }
  \label{fig:tmp4}
\end{figure}

We use the idea of nonequilibrium linear response to investigate the
nonequilibrium phase space probability density, denoted by $\rho^\eps(q, p, t)$, of a one-dimensional model system: one particle in a
splitting single-well potential as shown in Fig.~\ref{fig:tmp4}.  For convenience, we let the mass of
the particle to be 1~\textsf{amu}, and the friction coefficient to be
$1\,\textsf{ps}^{-1}$.
The temperature is the
room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
The unperturbed Hamiltonian of the system is
given by:
\begin{align}
  H ( p,  q) = \frac 12  p^2 + U( q) 
\end{align}
with potential
\begin{align}
  U( q) = \frac12\,k\, q^2 
\end{align}
Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^2)$.
See the red line in Fig.~\ref{fig:tmp4} for the  potential $U$.
%$\textsf{kJ} / (\textsf{mol nm}^4)}$
The nonequilibrium driving $ D$ is given by a force of gradient form:
\begin{align}
   D( q) = -\nabla_{ q} V( q) ,
\end{align}
where the driving potential $V( q)$ has a Gaussian profile:
\begin{align}
  V( q) = \frac{1}{\sqrt{2\pi \Sigma^2}}
  \exp\Big\{-\frac{ q^2}{2\Sigma^2}\Big\}
\end{align}
we use $\Sigma = 0.16$~\textsf{nm}.  The strength of nonequilibrium
driving $u(t)$ is set to be linearly growing, i.e. $ u(t) = k_e\cdot t/T$,
where $k_e$ is a unitless constant.
We
consider the perturbation to the system given by $\eps\delta u(t) = \eps k_e \cdot t/T$.
We use the following parameters: end time $T = 20$~\textsf{ps}, 
$k_e = 1$ and $\eps = 1$, see Fig.~\ref{fig:tmp4}
for the nonequilibrium driving potential and perturbed
potential at time $t = T$. The initial distribution $\rho^0(q, p, 0)$ is set to
be equilibrium distribution of the unperturbed system.
  We use Euler scheme with time step of $10^{-4}$ to discretize the
  Langevin equation. The initial configurations for nonequilibrium
  simulation are generated by taking configurations
  from an equilibrium simulation of $10^6$~\textsf{ps} at an 1~\textsf{ps} time interval.
  Then starting with these initial configurations according to $\rho_{\infty}\propto\exp(-\beta H)$,  the nonequilibrium
  Langevin equation is integrated until 20~\textsf{ps} with the same numerical
  scheme and time step.
